---
title: "Building the displacement kernel — a walkthrough"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

> **Before you start:** this file is meant to be stepped through chunk by chunk with
> the green "run current chunk" button (or Cmd/Ctrl+Enter), not Knit-ed top to bottom —
> several chunks loop over 30 day-partitions and write real files to `E:/`. Knitting
> is fine once everything downstream is stable and you want a rendered report.



A click-through build of the new pipeline, in order, with a short SQL lesson before
each chunk and checkpoints that print something you can look at. Exercise answers are
in the appendix.

Read the lesson, run the chunk, look at the checkpoint, do the exercise. If a
checkpoint looks wrong, stop there — every later stage inherits it.

---

## What we are building, in one paragraph

One table. For every device-day, a row per place the device was, carrying the distance
of that place from the device's home and a weight. Everything downstream — the
diagonal, the distance bands, the 250 km cutoff, the kernel — is a way of grouping
that one table. Nothing about bin edges or cutoffs gets baked in; they are all filters
applied at the end.

Stages: work out where each device lives (§1), put every visit on the map (§2),
measure how far each visit is from that device's home (§3), decide what each visit
weighs (§4), give every device one vote (§5), then bin (§6) and check (§7).

---

## Two decisions settled before we start

**Post-hoc binning: yes, and that's the point.** You asked whether we can decide later
what counts as "zero movement". We can, because `d_km` stays on the table as a
continuous number and never gets collapsed into a category. Choosing a home threshold
becomes one line in a `mutate()` at the very end, and you can produce the profile at
0.5 km, 2 km and 5 km side by side for the cost of running that line three times.

One caveat, and it's a real one. Post-hoc freedom does not create resolution that
isn't there. Where a visit's position came from a geohash-5 cell, the position is only
good to about 5 km, so a threshold below that is measuring cell-boundary luck rather
than behaviour. That's why §2 carries a `loc_src` flag — so you can restrict to
block-group-resolved rows when you want to look below 5 km, and report how much of the
sample that leaves you.

**Trip weighting versus dwell weighting.** Your point about the single visit is right,
and it's a genuine fork rather than a detail. A device-day with seven hours at home and
one twenty-minute stop at *j* is, under dwell weighting, 95% "at home" — the trip
barely registers. Under trip weighting it's a trip to *j*.

These answer different questions, and neither is wrong:

- **Dwell weighting** estimates *where people spend time*. It is the like-for-like
  match to Meta, whose `distance_category_ping_fraction` is a share of pings, which is
  a time share.
- **Trip weighting** estimates *where people go*. Arguably the more interesting
  quantity, and much less sensitive to the dwell censoring we found — but it is not
  what Meta reports, so an offset fitted on trip-weighted Veraset against
  ping-weighted Meta would absorb the definitional gap into the parameter.

So we carry both. The weight is a column, the choice is a switch at the end, and §7
shows you how far apart they land. My suggestion is to fit the Meta mapping on dwell
weighting and report trip weighting as the substantive result, but see the numbers
first.

**One user, one vote.** You confirmed Meta normalises by users, so we do the same: a
visit's weight is normalised within its device-day, device-days are averaged within
device, devices are averaged within county. §5 does this in three steps.

---

# §0 — Setup

## SQL lesson 1: what DuckDB is doing here

DuckDB is not a database you load data into. It's a query engine that reads parquet
files off disk where they lie. `read_parquet('E:/.../*.parquet')` is the table — the
files never move.

Two consequences worth internalising:

**Naming your columns is a performance decision.** Parquet stores each column
separately, so `SELECT caid, minimum_dwell FROM ...` physically reads two columns and
skips sixteen. `SELECT *` reads all eighteen. On `other_visits` that's most of the
difference between minutes and an afternoon.

**`DESCRIBE` is free.** `DESCRIBE SELECT * FROM read_parquet('...') LIMIT 0` reads
only the file footers. Use it before every new scan.

## SQL lesson 2: how to read a query

SQL is *written* with `SELECT` first but *evaluated* in roughly dplyr's order:

```
FROM  →  WHERE  →  GROUP BY  →  SELECT  →  HAVING  →  ORDER BY  →  LIMIT
```

So when you read a query, skip to `FROM`, then `WHERE`, then `GROUP BY`, and read the
`SELECT` list *last*. It's a `summarise()` that someone moved to the top of the pipe.

The translation table:

| dplyr | SQL |
|---|---|
| the data | `FROM read_parquet('...')` |
| `select()` | the column list after `SELECT` |
| `mutate()` | an expression in the `SELECT` list, named with `AS` |
| `filter()` before grouping | `WHERE` |
| `group_by()` | `GROUP BY` |
| `summarise()` | aggregate functions in the `SELECT` list |
| `filter()` after summarising | `HAVING` |
| `arrange()` | `ORDER BY` |
| `bind_rows()` | `UNION ALL` |
| piping into a named step | `WITH name AS (...)` |

And the one behavioural difference that will bite you: **NULL is not `NA`**. `x = NULL`
is never true — use `x IS NULL`. Aggregates *skip* NULLs silently, so `SUM(dwell)`
needs no `na.rm` and gives no warning about what it dropped.

```{r chunk-01}
library(DBI)
library(duckdb)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(glue)


# --- Paths -----------------------------------------------------------------

# DATA_ROOT is the one thing to change per machine. Every subtree below is
# derived from it: raw pulls live in raw/, external reference data (Meta,
# RUCC, crosswalks) in external/, everything this notebook builds in kernel/,
# and duckdb's spill space in tmp/. Every other script in this repo uses the
# same DATA_ROOT and the same subtree names.
DATA_ROOT <- "E:/dewey-data"
RAW_DIR      <- file.path(DATA_ROOT, "raw")
EXTERNAL_DIR <- file.path(DATA_ROOT, "external")
TMP_DIR      <- file.path(DATA_ROOT, "tmp")
OUT          <- file.path(DATA_ROOT, "kernel")
# Forward-slash form of TMP_DIR for embedding in DuckDB SQL strings, since
# backslashes are meaningful in SQL string literals on Windows paths.
TMP_DIR_SQL  <- gsub("\\\\", "/", TMP_DIR)

dir.create(file.path(OUT, "home_parts"),   showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(OUT, "visit_parts"),  showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(OUT, "devday_parts"), showWarnings = FALSE, recursive = TRUE)


# --- Settings --------------------------------------------------------------

SOURCES   <- c("home_visits", "work_visits", "other_visits")
MEM_LIMIT <- "24GB"
N_THREADS <- 32

CBG_AREA_MAX <- 25      # km^2; above this we prefer the geohash cell
TAU_MINUTES  <- 15      # weight given to a zero-dwell (single-ping) visit


# --- Connection ------------------------------------------------------------

con <- dbConnect(duckdb::duckdb(), dbdir = ":memory:")

dbExecute(con, glue("SET temp_directory = '{TMP_DIR_SQL}'"))
dbExecute(con, glue("SET memory_limit = '{MEM_LIMIT}'"))
dbExecute(con, glue("SET threads = {N_THREADS}"))
dbExecute(con, "SET preserve_insertion_order = false")


# --- Two small helpers, defined once at the top ----------------------------

day_glob <- function(src, day) {
  glue("{RAW_DIR}/{src}/{day}/*.parquet")
}

list_days <- function(src) {
  sort(list.dirs(file.path(RAW_DIR, src), recursive = FALSE, full.names = FALSE))
}
```
```{r}

```

## Checkpoint 0b — how many threads actually help

64 physical cores, 128 logical (hyperthreaded). The hash-aggregation stages in this
pipeline (§1, §5) are memory-bound, pointer-chasing work — the two logical threads on
one physical core share cache and execution ports, so the hyperthread sibling often
adds little to that kind of workload. Rather than assume, benchmark it on one real day
of `other_visits` before picking `N_THREADS` for the full month.

Memory is not the constraint on this machine — with hundreds of GB free, set
`MEM_LIMIT` generously and hold it fixed across the whole benchmark so you're reading
off thread count alone, not confounding it with memory pressure.

```{r chunk-00b, eval=FALSE}
BENCH_DAY     <- list_days("other_visits")[1]
THREAD_GRID   <- c(8, 16, 32, 64, 96, 128)
BENCH_MEM_GB  <- "400GB"   # generous and fixed across every run

bench_sql <- glue("
  SELECT
    hash(caid)                                                        AS did,
    substr(lpad(CAST(census_block_group AS VARCHAR), 12, '0'), 1, 12) AS cbg,
    SUM(GREATEST(TRY_CAST(minimum_dwell AS DOUBLE), 0))                AS dwell
  FROM read_parquet('{day_glob(\"other_visits\", BENCH_DAY)}')
  WHERE census_block_group IS NOT NULL
  GROUP BY did, cbg
")

# One fresh connection per setting, so no state or cache carries over.
thread_bench <- map_dfr(THREAD_GRID, function(n_threads) {
  bench_con <- dbConnect(duckdb::duckdb(), dbdir = ":memory:")
  dbExecute(bench_con, glue("SET temp_directory = '{TMP_DIR_SQL}'"))
  dbExecute(bench_con, glue("SET memory_limit = '{BENCH_MEM_GB}'"))
  dbExecute(bench_con, glue("SET threads = {n_threads}"))
  dbExecute(bench_con, "SET preserve_insertion_order = false")

  elapsed <- system.time(dbGetQuery(bench_con, bench_sql))[["elapsed"]]
  dbDisconnect(bench_con, shutdown = TRUE)

  tibble(n_threads = n_threads, seconds = elapsed)
})

thread_bench |>
  mutate(speedup = seconds[n_threads == min(n_threads)] / seconds)
```

Read it as a curve, not a single number: look for where `seconds` stops dropping.
Also worth a glance — `DESCRIBE` won't tell you this, but row-group count caps how
finely a scan can be split. If `other_visits` files are written with only a few
hundred row groups per day, that's a second ceiling independent of thread count:

```{r chunk-00c, eval=FALSE}
dbGetQuery(con, glue("
  SELECT file_name, COUNT(*) AS n_row_groups
  FROM parquet_metadata('{day_glob(\"other_visits\", BENCH_DAY)}')
  GROUP BY file_name
"))
```

> **Exercise 0b.** If the curve in `thread_bench` flattens at, say, 32 threads, is
> that evidence the workload is memory-bandwidth-bound, row-group-limited, or both?
> What second query from this section would you run to tell them apart?
### Checkpoint 0

```{r chunk-02}
days <- list_days("home_visits")

tibble(src = SOURCES) |>
  mutate(n_days = map_int(src, ~ length(list_days(.x))))
```

Expect 30 for each. If the day names aren't `2025-06-01` style, everything downstream
still works — `day_glob()` just pastes whatever `list_days()` returns.

> **Exercise 0.** Run `DESCRIBE SELECT * FROM read_parquet(...) LIMIT 0` against one
> day of `other_visits` and confirm `minimum_dwell` comes back as `DOUBLE`. Then time
> it. It should be effectively instant on 240M rows — convince yourself the footer
> claim is true.

---

# §1 — Where each device lives

## SQL lesson 3: `GROUP BY`, and picking the argmax without a window function

We want, per device, the block group where it accumulated the most home dwell. In
dplyr you'd `group_by(caid, cbg) |> summarise(d = sum(dwell)) |> slice_max(d)`.

The first half is a plain `GROUP BY` on two keys. The second half — "which group had
the max" — is where people reach for a window function (`ROW_NUMBER() OVER (PARTITION
BY ...)`), and your notes are right that window functions over huge tables are
dangerous, because they need a full sort and sorts spill worse than aggregates.

DuckDB gives you `arg_max(a, b)`: the value of `a` at the row where `b` is largest.
It's an ordinary aggregate, so it streams.

```sql
SELECT
  did,
  arg_max(cbg, dwell) AS home_cbg,   -- the cbg with the most dwell
  MAX(dwell)          AS home_dwell, -- how much dwell that was
  SUM(dwell)          AS total_dwell -- across all cbgs
FROM per_device_cbg
GROUP BY did
```

The ratio `home_dwell / total_dwell` is the concentration flag — a device 95%
concentrated at one address is a different animal from one split evenly across three,
and that ratio is what should catch the Shelby County long-haul drivers.

Two other things in this chunk you've met before: `hash(caid)` turns a 36-character
string into an 8-byte integer that sorts and spills far cheaper, and the day loop with
`.tmp`-then-rename means a crash costs one day rather than everything.

_This chunk loops over all 30 days and writes to `E:/`. Set `eval=FALSE` above so a "Run All" or Knit can't trigger it by accident — run it deliberately with the green triangle._

```{r chunk-03, eval=FALSE}
# --- 1a. Per-day partial sums: device x block group ------------------------

home_day_sql <- function(day) {
  glue("
    SELECT
      hash(caid)                                                        AS did,
      substr(lpad(CAST(census_block_group AS VARCHAR), 12, '0'), 1, 12) AS cbg,
      SUM(GREATEST(TRY_CAST(minimum_dwell AS DOUBLE), 0))               AS dwell,
      COUNT(*)                                                          AS n_rows
    FROM read_parquet('{day_glob(\"home_visits\", day)}')
    WHERE census_block_group IS NOT NULL
    GROUP BY did, cbg
  ")
}

for (day in list_days("home_visits")) {
  out_path <- file.path(OUT, "home_parts", glue("{day}.parquet"))
  if (file.exists(out_path)) next

  message(glue("home {day} ..."))
  tmp_path <- paste0(out_path, ".tmp")

  dbExecute(con, glue("
    COPY ({home_day_sql(day)}) TO '{tmp_path}' (FORMAT PARQUET)
  "))

  file.rename(tmp_path, out_path)
}


# --- 1b. Roll the month up into one home assignment ------------------------

home_parts_glob <- glue("{OUT}/home_parts/*.parquet")

dbExecute(con, glue("
  COPY (
    WITH monthly AS (
      SELECT
        did,
        cbg,
        SUM(dwell)  AS dwell,
        SUM(n_rows) AS n_rows
      FROM read_parquet('{home_parts_glob}')
      GROUP BY did, cbg
    )
    SELECT
      did,
      arg_max(cbg, dwell)  AS home_cbg,
      MAX(dwell)           AS home_dwell,
      SUM(dwell)           AS total_home_dwell,
      SUM(n_rows)          AS n_home_rows,
      COUNT(*)             AS n_home_cbgs
    FROM monthly
    GROUP BY did
  ) TO '{OUT}/home_assignment.parquet.tmp' (FORMAT PARQUET)
"))

file.rename(
  file.path(OUT, "home_assignment.parquet.tmp"),
  file.path(OUT, "home_assignment.parquet")
)
```

A wrinkle worth noticing. `arg_max(cbg, dwell)` breaks ties arbitrarily, and after the
dwell audit we know a third of home rows have zero dwell. A device whose home dwell is
*all* zero gets an essentially random block group. Count those and decide what to do
with them — the checkpoint below surfaces the number.

### Checkpoint 1

```{r chunk-04}
home_summary <- dbGetQuery(con, glue("
  SELECT
    COUNT(*)                                                     AS n_devices,
    SUM(CASE WHEN total_home_dwell = 0 THEN 1 ELSE 0 END)        AS n_zero_dwell,
    SUM(CASE WHEN n_home_cbgs = 1 THEN 1 ELSE 0 END)             AS n_single_cbg,
    AVG(n_home_cbgs)                                             AS mean_cbgs,
    approx_quantile(home_dwell / NULLIF(total_home_dwell, 0), 0.10) AS conc_p10,
    approx_quantile(home_dwell / NULLIF(total_home_dwell, 0), 0.50) AS conc_p50
  FROM read_parquet('{OUT}/home_assignment.parquet')
")) |>
  as_tibble()

home_summary |>
  mutate(
    zero_dwell_share = n_zero_dwell / n_devices,
    single_cbg_share = n_single_cbg / n_devices
  )
```

`n_devices` should land near 42.5M — that's the home panel from your coverage audit,
and it's the check that this stage reproduces what you already know. `NULLIF(x, 0)`
turns a zero denominator into NULL rather than an error, which is the SQL idiom for
"don't divide by zero".

> **Exercise 1a.** What share of devices have their home dwell spread across more than
> three block groups? Write it as one `dbGetQuery()` using `SUM(CASE WHEN ... THEN 1
> ELSE 0 END)`.
>
> **Exercise 1b.** Rewrite the 1b query using `MAX(dwell)` and a join back to
> `monthly` instead of `arg_max`. Then say in one sentence why `arg_max` is preferable
> at this scale.
>
> **Exercise 1c.** The concentration ratio uses dwell. Build the alternative using
> `n_rows` instead, and check whether the two disagree for many devices. If they do,
> which would you trust given what we learned about single-ping visits?

---
```{r}
try_1 <- dbGetQuery(con, glue("
SELECT
COUNT(*) AS n_devices,
SUM(CASE WHEN total_home_dwell = 0 THEN 1 ELSE 0 END) AS n_zero_dwell,
SUM(CASE WHEN n_home_cbgs = 1 THEN 1 ELSE 0 END) as n_trip,
  FROM read_parquet('{OUT}/home_assignment.parquet')
")) |>
  as_tibble()


##exercise B 


dbExecute(con, glue("
  COPY (
    WITH monthly AS (
      SELECT
        did,
        cbg,
        SUM(dwell)  AS dwell,
        SUM(n_rows) AS n_rows
      FROM read_parquet('{home_parts_glob}')
      GROUP BY did, cbg
    )
    SELECT
      did,
      arg_max(cbg, dwell)  AS home_cbg,
      MAX(dwell)           AS home_dwell,
      SUM(dwell)           AS total_home_dwell,
      SUM(n_rows)          AS n_home_rows,
      COUNT(*)             AS n_home_cbgs
    FROM monthly
    GROUP BY did
  ) 
"))

```


# §2 — Putting every visit on the map

## SQL lesson 4: knowing when to leave the database

The rule for this project: SQL for anything that touches the billion-row scans,
R for anything small. The geography lookup is small — a few hundred thousand rows —
so we build it in dplyr and *hand it to* DuckDB with `dbWriteTable()`, which copies an
R data frame in as a real table the SQL can join against.

(`duckdb_register()` does the same thing without copying, as a view over the R object.
Fine for a one-off query; `dbWriteTable()` is better here because we join against it
inside a thirty-iteration loop.)

Geohash decoding is pure arithmetic and belongs in R. A geohash is base-32: each
character carries 5 bits, and the bits alternate longitude, latitude, longitude, …
starting with longitude. Five characters is 25 bits — 13 for longitude, 12 for
latitude — which is why the cells are about 4.9 km square at the equator and narrower
as you go north.

```{r chunk-05}
GEOHASH_ALPHABET <- str_split("0123456789bcdefghjkmnpqrstuvwxyz", "")[[1]]

# The one place in this script that uses matrix indexing. Bit arithmetic has no
# tidy idiom and wrapping it would obscure it.
decode_geohash5 <- function(geohash) {
  digits <- str_split_fixed(str_to_lower(geohash), "", 5)
  values <- matrix(match(digits, GEOHASH_ALPHABET) - 1, ncol = 5)

  packed <- values[, 1] * 32^4 + values[, 2] * 32^3 + values[, 3] * 32^2 +
            values[, 4] * 32   + values[, 5]

  lon_bits <- 0
  lat_bits <- 0

  for (i in 1:25) {
    bit <- (packed %/% 2^(25 - i)) %% 2
    if (i %% 2 == 1) {
      lon_bits <- lon_bits * 2 + bit
    } else {
      lat_bits <- lat_bits * 2 + bit
    }
  }

  tibble(
    geohash_5 = geohash,
    gh_lon    = -180 + (lon_bits + 0.5) * 360 / 2^13,
    gh_lat    =  -90 + (lat_bits + 0.5) * 180 / 2^12
  )
}
```

Now the block group centroids. You need the 2020 Census block group gazetteer or the
TIGER shapefiles; the columns we want are the 12-digit GEOID, latitude, longitude and
land area.

```{r chunk-06}

cbg_centroids <- read_csv(file.path(EXTERNAL_DIR, "cbg_centroids.csv"),
                          col_types = cols(cbg = col_character())) |>
  select(cbg, cbg_lat, cbg_lon, area_km2) |>
  filter(!is.na(cbg_lat), !is.na(cbg_lon))


dbWriteTable(con, "cbg_lookup", cbg_centroids, overwrite = TRUE)


```

And the distinct geohashes actually present in the data — pull the list out of SQL,
decode in R, push the decoded table back.

```{r chunk-07}

geohash_list <- dbGetQuery(con, glue("
  SELECT DISTINCT geohash_5
  FROM read_parquet([
    '{RAW_DIR}/home_visits/*/*.parquet',
    '{RAW_DIR}/work_visits/*/*.parquet',
    '{RAW_DIR}/other_visits/*/*.parquet'
  ], union_by_name = true)
  WHERE geohash_5 IS NOT NULL
")) |>
  as_tibble()

geohash_lookup <- decode_geohash5(geohash_list$geohash_5)

dbWriteTable(con, "geohash_lookup", geohash_lookup, overwrite = TRUE)
```

That `DISTINCT` over the whole month is one of the few places a distinct is safe: it's
a single ungrouped distinct over one narrow column, not a distinct *per group* across
hundreds of millions of groups. Different animal entirely — the thing that killed
`other_visits` was needing a separate hash set for every one of half a billion groups.

### Checkpoint 2

```{r chunk-08}
geohash_lookup |>
  summarise(
    n_cells = n(),
    lat_min = min(gh_lat), lat_max = max(gh_lat),
    lon_min = min(gh_lon), lon_max = max(gh_lon)
  )

geohash_lookup |>
  summarise(
    n_positive_lon = sum(gh_lon > 0),
    n_southern     = sum(gh_lat < 0),
    n_far_north    = sum(gh_lat > 60)
  )
```

Expect a few hundred thousand cells, latitudes roughly 18–72 (Puerto Rico to Alaska)
and longitudes roughly -180 to -65. Anything outside that means the decoder is wrong.

The better validation is self-consistency: a row's geohash centre and its block group
centroid describe the same visit, so they should agree to within the coarser of the two
resolutions.

```{r chunk-09}
agreement <- dbGetQuery(con, glue("
  WITH sample AS (
    SELECT
      substr(lpad(CAST(census_block_group AS VARCHAR), 12, '0'), 1, 12) AS cbg,
      geohash_5
    FROM read_parquet('{day_glob(\"other_visits\", \"2025-06-03\")}')
    WHERE hash(caid) % 1000 = 0
      AND census_block_group IS NOT NULL
      AND geohash_5 IS NOT NULL
  )
  SELECT
    c.area_km2,
    2 * 6371.0088 * asin(sqrt(
      pow(sin(radians(g.gh_lat - c.cbg_lat) / 2), 2) +
      cos(radians(c.cbg_lat)) * cos(radians(g.gh_lat)) *
      pow(sin(radians(g.gh_lon - c.cbg_lon) / 2), 2)
    )) AS gap_km
  FROM sample s
  JOIN cbg_lookup     c ON s.cbg       = c.cbg
  JOIN geohash_lookup g ON s.geohash_5 = g.geohash_5
")) |>
  as_tibble()

agreement |>
  mutate(size_class = if_else(area_km2 <= CBG_AREA_MAX, "small cbg", "large cbg")) |>
  group_by(size_class) |>
  summarise(
    n      = n(),
    p50    = median(gap_km),
    p90    = quantile(gap_km, 0.9),
    p99    = quantile(gap_km, 0.99),
    .groups = "drop"
  )
```

For small block groups the gap should mostly be under ~3.5 km, which is the worst case
within a geohash cell. For large ones it can be much bigger, and that's the whole
argument for the size rule. If small block groups show large gaps, the join is
misaligned — check leading zeros first, since that's the classic failure.

> **Exercise 2a.** Decode `9q8yy` by hand-ish: run `decode_geohash5("9q8yy")` and check
> the result is in the San Francisco Bay Area. Then decode `dhwux`, which appears in
> your home sample, and confirm it lands in south Florida.
>
> **Exercise 2b.** The `JOIN`s above are inner joins, so rows with no matching block
> group vanish silently. Change one to a `LEFT JOIN` and count how many rows have no
> centroid. Should that number worry us?
>
> **Exercise 2c.** `CBG_AREA_MAX` is set to 25 km². Recompute the agreement table at
> 5, 25 and 100 and argue for a value.

---

# §3 — How far each visit is from home

## SQL lesson 5: joins, and why we don't build a pair table

A `JOIN` matches rows between two tables on a key. `INNER JOIN` keeps only matches;
`LEFT JOIN` keeps every row of the left table, filling NULL where there's no match.
This is `inner_join()` and `left_join()`, same semantics.

The temptation at this point is to precompute a distance table for every pair of block
groups. Don't: 240,000 block groups is 29 billion pairs. Instead we join the home
coordinates and the visit coordinates onto each row and compute the distance *inline*.
Trigonometry on a row is nanoseconds; a 29-billion-row lookup is a research project.

This also keeps distance continuous and per-row, which is precisely what makes the
post-hoc binning in §6 possible.

The location rule from §2 becomes a `COALESCE` with a `CASE`: use the block group
centroid when the block group is small, otherwise the geohash centre, otherwise
whichever exists.

_Same as above — 30-day loop, writes to disk. Run deliberately._

```{r chunk-10, eval=FALSE}
visit_day_sql <- function(day) {
  home_glob  <- day_glob("home_visits",  day)
  work_glob  <- day_glob("work_visits",  day)
  other_glob <- day_glob("other_visits", day)

  glue("
    WITH stacked AS (
      SELECT hash(caid) AS did, utc_timestamp, census_block_group, geohash_5,
             TRY_CAST(minimum_dwell AS DOUBLE) AS dwell, 1 AS is_home
      FROM read_parquet('{home_glob}')
      UNION ALL
      SELECT hash(caid), utc_timestamp, census_block_group, geohash_5,
             TRY_CAST(minimum_dwell AS DOUBLE), 0
      FROM read_parquet('{work_glob}')
      UNION ALL
      SELECT hash(caid), utc_timestamp, census_block_group, geohash_5,
             TRY_CAST(minimum_dwell AS DOUBLE), 0
      FROM read_parquet('{other_glob}')
    ),

    tagged AS (
      SELECT
        s.did,
        CAST(CAST(to_timestamp(s.utc_timestamp) AS TIMESTAMP) AS DATE) AS day,
        substr(lpad(CAST(s.census_block_group AS VARCHAR), 12, '0'), 1, 12) AS cbg,
        s.geohash_5,
        s.dwell,
        s.is_home
      FROM stacked s
    ),

    placed AS (
      SELECT
        t.did,
        t.day,
        t.cbg,
        substr(t.cbg, 1, 5) AS county,
        t.dwell,
        t.is_home,
        CASE WHEN c.cbg IS NOT NULL AND c.area_km2 <= {CBG_AREA_MAX}
             THEN 'cbg' ELSE 'geohash' END AS loc_src,
        COALESCE(
          CASE WHEN c.area_km2 <= {CBG_AREA_MAX} THEN c.cbg_lat END,
          g.gh_lat, c.cbg_lat
        ) AS lat,
        COALESCE(
          CASE WHEN c.area_km2 <= {CBG_AREA_MAX} THEN c.cbg_lon END,
          g.gh_lon, c.cbg_lon
        ) AS lon
      FROM tagged t
      LEFT JOIN cbg_lookup     c ON t.cbg       = c.cbg
      LEFT JOIN geohash_lookup g ON t.geohash_5 = g.geohash_5
    )

    SELECT
      p.did,
      p.day,
      p.cbg,
      p.county,
      p.loc_src,
      p.is_home,
      p.dwell,
      2 * 6371.0088 * asin(sqrt(
        pow(sin(radians(p.lat - hc.cbg_lat) / 2), 2) +
        cos(radians(hc.cbg_lat)) * cos(radians(p.lat)) *
        pow(sin(radians(p.lon - hc.cbg_lon) / 2), 2)
      )) AS d_km
    FROM placed p
    JOIN read_parquet('{OUT}/home_assignment.parquet') h ON p.did = h.did
    JOIN cbg_lookup hc ON h.home_cbg = hc.cbg
    WHERE p.lat IS NOT NULL
  ")
}

for (day in list_days("home_visits")) {
  out_path <- file.path(OUT, "visit_parts", glue("{day}.parquet"))
  if (file.exists(out_path)) next

  message(glue("visits {day} ..."))
  tmp_path <- paste0(out_path, ".tmp")

  dbExecute(con, glue("COPY ({visit_day_sql(day)}) TO '{tmp_path}' (FORMAT PARQUET)"))
  file.rename(tmp_path, out_path)
}
```

The `JOIN` against `home_assignment` is doing double duty: it attaches the home
coordinates *and* restricts to the home panel, because a device with no home row has no
matching row and drops out. That's the semi-join-inside-the-scan pattern from your
performance notes, and it's why this stage is tractable — it discards 91.5% of devices
during the scan rather than after it.

### Checkpoint 3

```{r chunk-11}
one_day <- dbGetQuery(con, glue("
  SELECT
    loc_src,
    is_home,
    COUNT(*)                                    AS n,
    approx_quantile(d_km, 0.50)                 AS p50_km,
    approx_quantile(d_km, 0.90)                 AS p90_km,
    SUM(CASE WHEN d_km < 1   THEN 1 ELSE 0 END) AS n_under_1km,
    SUM(CASE WHEN d_km > 250 THEN 1 ELSE 0 END) AS n_over_250km
  FROM read_parquet('{OUT}/visit_parts/2025-06-03.parquet')
  GROUP BY loc_src, is_home
")) |>
  as_tibble()

one_day |>
  mutate(
    under_1km_share = n_under_1km / n,
    over_250km_share = n_over_250km / n
  ) |>
  arrange(desc(is_home), loc_src)
```

What to look for: home rows (`is_home = 1`) should be overwhelmingly at `d_km` near
zero, because a device's home visit should be at its own home block group. If a large
fraction of home rows sit far from home, the modal assignment in §1 is picking the
wrong place and nothing downstream will make sense. This is the single most diagnostic
number in the whole build.

> **Exercise 3a.** What share of home rows have `d_km > 50`? Given that these are
> devices whose *own* home visits are 50 km from their assigned home, propose a rule
> for what to do with them.
>
> **Exercise 3b.** The haversine expression appears twice in this document. Rather
> than repeat it, DuckDB lets you define a macro:
> `CREATE MACRO hav(lat1, lon1, lat2, lon2) AS ...`. Write it and substitute it in.
>
> **Exercise 3c.** Why does `WHERE p.lat IS NOT NULL` appear at the end rather than
> inside `placed`? Would putting it earlier change the answer, or only the speed?

---

# §4 — What each visit weighs

## SQL lesson 6: carrying alternatives rather than choosing

Nothing new syntactically — this is `CASE WHEN` again. The idea is the discipline:
every time you face a weighting choice, add a column instead of making the choice.
Storage isn't a constraint here and the alternative is re-running §3.

Three weights per row:

- `w_dwell` — dwell in minutes, with zero-dwell (single-ping) visits given `TAU_MINUTES`
  rather than nothing, and negatives clipped. Recall that dwell is last-ping minus
  first-ping, so every value is a lower bound and zeros mean "seen once", not "stayed
  no time".
- `w_trip` — one per visit, home included.
- `w_trip_nonhome` — one per visit, home excluded.

```{r chunk-12}
weight_sql <- function(day) {
  glue("
    SELECT
      did, day, cbg, county, loc_src, is_home, d_km, dwell,
      CASE
        WHEN dwell IS NULL  THEN {TAU_MINUTES}
        WHEN dwell <  0     THEN {TAU_MINUTES}
        WHEN dwell =  0     THEN {TAU_MINUTES}
        ELSE dwell
      END                                      AS w_dwell,
      1                                        AS w_trip,
      CASE WHEN is_home = 1 THEN 0 ELSE 1 END  AS w_trip_nonhome
    FROM read_parquet('{OUT}/visit_parts/{day}.parquet')
  ")
}
```

### Checkpoint 4

```{r chunk-13}
weight_check <- dbGetQuery(con, glue("
  SELECT
    is_home,
    COUNT(*)                    AS n,
    SUM(w_dwell)                AS total_dwell,
    AVG(w_dwell)                AS mean_dwell,
    approx_quantile(w_dwell, 0.5) AS p50_dwell
  FROM ({weight_sql('2025-06-03')})
  GROUP BY is_home
")) |>
  as_tibble()

weight_check |>
  mutate(share_of_weight = total_dwell / sum(total_dwell))
```

The number to stare at is `share_of_weight` for home. Under dwell weighting home is
going to take most of it — median home dwell was 411 minutes per device-day against
about 20 minutes for a typical other visit. If home comes out above ~90% of total
weight, that's your warning that the dwell-weighted diagonal will be very high and the
trip weighting is doing the substantive work.

> **Exercise 4a.** Recompute the table at `TAU_MINUTES` of 1, 15 and 60. How much does
> home's share of weight move? That range is your sensitivity, and it should go in the
> write-up.
>
> **Exercise 4b.** The `CASE` has three branches that all return `TAU_MINUTES`.
> Collapse them into one condition. (Careful with NULL — `dwell < 0` is not TRUE when
> `dwell` is NULL, and `COALESCE` is your friend.)
>
> **Exercise 4c.** Argue for or against a fourth weight: dwell, but with home capped at
> some ceiling to stop it swamping everything. What would that quantity mean?

---

# §5 — One user, one vote

## SQL lesson 7: grouping twice instead of counting distinct

Meta normalises by user, so we do too, in three steps:

1. within a device-day, weights sum to 1;
2. within a device, device-days average to 1;
3. within a county, devices average to 1.

Step 1 needs each row's weight divided by its device-day total. In dplyr that's
`group_by(did, day) |> mutate(share = w / sum(w))`. In SQL the direct translation is a
window function, which we're avoiding — so instead we aggregate to get the totals and
join them back. Two passes, both streaming, no sort.

This is also where the distinct-destination count comes from, and it's the pattern
worth remembering: **group to the finer key first, then count rows at the coarser key.**
The first grouping deduplicates as a side effect, so the second only has to count. That
gets you a distinct count without `COUNT(DISTINCT)`, which is what stalled
`other_visits` at 10%.

_Same as above — 30-day loop, writes to disk. Run deliberately._

```{r chunk-14, eval=FALSE}
devday_sql <- function(day) {
  glue("
    WITH w AS ({weight_sql(day)}),

    -- collapse to one row per device-day per location
    loc AS (
      SELECT
        did, day, cbg, county, loc_src,
        MAX(is_home)         AS is_home,
        MIN(d_km)            AS d_km,
        SUM(w_dwell)         AS w_dwell,
        SUM(w_trip)          AS w_trip,
        SUM(w_trip_nonhome)  AS w_trip_nonhome
      FROM w
      GROUP BY did, day, cbg, county, loc_src
    ),

    -- device-day totals (COUNT(*) here IS the distinct location count)
    totals AS (
      SELECT
        did, day,
        COUNT(*)                AS n_locations,
        SUM(w_dwell)            AS tot_dwell,
        SUM(w_trip)             AS tot_trip,
        SUM(w_trip_nonhome)     AS tot_trip_nonhome
      FROM loc
      GROUP BY did, day
    )

    SELECT
      l.did, l.day, l.cbg, l.county, l.loc_src, l.is_home, l.d_km,
      t.n_locations,
      l.w_dwell        / t.tot_dwell                       AS s_dwell,
      l.w_trip         / t.tot_trip                        AS s_trip,
      CASE WHEN t.tot_trip_nonhome > 0
           THEN l.w_trip_nonhome / t.tot_trip_nonhome END  AS s_trip_nonhome
    FROM loc l
    JOIN totals t ON l.did = t.did AND l.day = t.day
  ")
}

for (day in list_days("home_visits")) {
  out_path <- file.path(OUT, "devday_parts", glue("{day}.parquet"))
  if (file.exists(out_path)) next

  message(glue("devday {day} ..."))
  tmp_path <- paste0(out_path, ".tmp")

  dbExecute(con, glue("COPY ({devday_sql(day)}) TO '{tmp_path}' (FORMAT PARQUET)"))
  file.rename(tmp_path, out_path)
}
```

`s_trip_nonhome` is NULL for home-only device-days — the `CASE` with no `ELSE` gives
NULL, aggregates skip it, and those device-days simply don't contribute to that
variant. That's the honest behaviour: a device-day with no observed trip has no opinion
about trip distances.

Steps 2 and 3 — averaging device-days within device, then devices within county — are
small enough to do after binning, in dplyr, in §6.

### Checkpoint 5

```{r chunk-15}
devday_check <- dbGetQuery(con, glue("
  SELECT
    n_locations,
    COUNT(*) AS n_rows
  FROM read_parquet('{OUT}/devday_parts/2025-06-03.parquet')
  GROUP BY n_locations
")) |>
  as_tibble()

devday_check |>
  arrange(n_locations) |>
  mutate(share = n_rows / sum(n_rows))
```

`n_locations = 1` should be a large share, but *smaller* than the 87% single-destination
figure from the fallback test — because that was counting distinct counties and this
counts distinct block groups, which is finer. If it isn't smaller, the location
resolution isn't buying us anything and that's worth understanding before going on.

Also worth a direct check that the normalisation worked:

```{r chunk-16}
dbGetQuery(con, glue("
  WITH sums AS (
    SELECT did, day, SUM(s_dwell) AS total
    FROM read_parquet('{OUT}/devday_parts/2025-06-03.parquet')
    GROUP BY did, day
  )
  SELECT MIN(total) AS min_total, MAX(total) AS max_total FROM sums
"))
```

Both should be 1 to within floating-point noise.

> **Exercise 5a.** `loc` groups by `cbg`, so two visits to different POIs in the same
> block group merge into one location. Is that right? What would change if we grouped
> by `safegraph_place_id` instead?
>
> **Exercise 5b.** `MIN(d_km)` in `loc` should be redundant — every row with the same
> `cbg` and `loc_src` has the same distance. Verify that with a query, and say what a
> violation would imply.
>
> **Exercise 5c.** Write the step-2 aggregation (device-days averaged within device) in
> dplyr, given a binned table with columns `did`, `day`, `band`, `share`.

---

# §6 — Binning, at the very end

Nothing here is SQL. The device-day table is now small enough per day to work with in
dplyr, and this is deliberately the stage where all the arbitrary choices live.

```{r chunk-17}
# Read the month, restricted to the ten benchmark origins to keep it quick.
selected <- read_csv(file.path(OUT, "narrow10", "selected_origins.csv"),
                     col_types = cols(fips = col_character()))

devday <- list.files(file.path(OUT, "devday_parts"), full.names = TRUE) |>
  map(~ dbGetQuery(con, glue("SELECT * FROM read_parquet('{.x}')"))) |>
  bind_rows() |>
  as_tibble()

home_county <- devday |>
  filter(is_home == 1) |>
  distinct(did, day, home_county = county)

profiles <- devday |>
  inner_join(home_county, by = c("did", "day")) |>
  filter(home_county %in% selected$fips)
```

Now the choices, all in one place, all reversible:

```{r chunk-18}
HOME_KM   <- 2      # what counts as "at home"
BAND_EDGE <- c(0, 1, 2.5, 5, 10, 25, 50, 100, 250, 500, 1000, Inf)

banded <- profiles |>
  mutate(
    at_home = d_km <= HOME_KM,
    band    = cut(d_km, breaks = BAND_EDGE, right = FALSE, include.lowest = TRUE)
  )

# step 2: device-days averaged within device
per_device <- banded |>
  group_by(home_county, did, band) |>
  summarise(share = sum(s_dwell) / n_distinct(day), .groups = "drop")

# step 3: devices averaged within county
county_profile <- per_device |>
  group_by(home_county, band) |>
  summarise(share = mean(share), n_devices = n_distinct(did), .groups = "drop") |>
  group_by(home_county) |>
  mutate(share = share / sum(share)) |>
  ungroup()
```

### Checkpoint 6

```{r chunk-19}
county_profile |>
  filter(home_county %in% c("06037", "30109", "19165")) |>
  select(home_county, band, share) |>
  pivot_wider(names_from = home_county, values_from = share) |>
  print(n = 20)

device_days <- banded |>
  distinct(home_county, did, day) |>
  count(home_county, did, name = "n_days")

per_device <- banded |>
  group_by(home_county, did, band) |>
  summarise(mass = sum(s_dwell), .groups = "drop") |>
  left_join(device_days, by = c("home_county", "did")) |>
  mutate(share = mass / n_days)


```

Los Angeles, Wibaux and Shelby side by side. LA should have a fat near-home mass and
almost nothing beyond 250 km; Wibaux should show mass at 50–250 km that LA doesn't,
because in Wibaux the shops are that far away. If Wibaux looks like LA, the home
assignment or the distance is wrong.

> **Exercise 6a.** Produce `county_profile` at `HOME_KM` of 0.5, 2 and 5 and put the
> three "at home" shares side by side. How much of the headline number is the threshold?
>
> **Exercise 6b.** Swap `s_dwell` for `s_trip` and rerun. Which counties move most, and
> is there a pattern by RUCC?
>
> **Exercise 6c.** Restrict to `loc_src == "cbg"` and rerun at `HOME_KM = 0.5`. How
> much of the sample survives, and is the sub-kilometre band now believable?

---

# §7 — Things to check before believing any of it

Six diagnostics, roughly in order of how likely they are to catch something.

1. **Home rows far from home.** From checkpoint 3. If more than a percent or two of
   home visits are far from the assigned home block group, §1 is misassigning.
2. **The 1000+ km bump.** Split the far tail by `n_locations` and by whether the
   device-day's weights were all τ. If the bump concentrates in single-ping device-days,
   it was the fallback all along.
3. **τ sensitivity.** The whole profile at τ = 1, 15, 60. Report the range, not a point.
4. **Dwell versus trip.** The same profile under `s_dwell` and `s_trip`. If they agree,
   the weighting debate is moot and you can say so. If they diverge, that divergence is
   a finding.
5. **Day trend.** `other_visits` coverage grows through June, so anything daily inherits
   a trend. Compute the profile for the first and last weeks separately; if they differ,
   pool the month and say why.
6. **Effective sample size.** About thirteen device-days per device, so the independent
   unit is the device. `n_devices` in `county_profile` is your real n, and Wibaux's will
   be small.
   
## Writing the profile out

`county_profile` is the handoff to Track B — the Stan fits and the Meta comparison run
on the analysis machine, so this needs to exist as a file, not just an object in the
session.

Two things travel with it. **The settings**, because a profile is meaningless without
knowing which `HOME_KM`, `TAU_MINUTES` and weighting produced it, and you will produce
several. **The device counts**, because Wibaux's profile rests on far fewer devices
than LA's and the Stan fit needs to know that — remember the independent unit is the
device, not the device-day.

```{r chunk-19b}
RUN_TAG <- glue("home{HOME_KM}km_tau{TAU_MINUTES}_dwell")

profile_out <- county_profile |>
  mutate(
    run_tag     = RUN_TAG,
    home_km     = HOME_KM,
    tau_minutes = TAU_MINUTES,
    weighting   = "s_dwell",
    built_on    = as.character(Sys.Date())
  ) |>
  relocate(run_tag, home_county, band, share, n_devices)

out_path <- file.path(OUT, glue("county_profile_{RUN_TAG}.csv"))
tmp_path <- paste0(out_path, ".tmp")

write_csv(profile_out, tmp_path)
file.rename(tmp_path, out_path)

out_path
```

The `RUN_TAG` in the filename is what stops you overwriting the dwell-weighted profile
with the trip-weighted one and then wondering months later which is which. Rerun §6
with different settings and each lands in its own file.

Worth also saving the band edges, since `cut()` labels are a fragile way to recover
them and Stan needs the numeric edges to condition the likelihood on the observed
range:

```{r chunk-19c}
tibble(
  run_tag   = RUN_TAG,
  band      = levels(cut(0, breaks = BAND_EDGE, right = FALSE, include.lowest = TRUE)),
  edge_low  = head(BAND_EDGE, -1),
  edge_high = tail(BAND_EDGE, -1)
) |>
  write_csv(file.path(OUT, glue("band_edges_{RUN_TAG}.csv")))
```

> **Exercise 6d.** Wrap §6 in a loop over `HOME_KM` of 0.5, 2 and 5, writing one file
> per setting. Then read all three back with `list.files()` and `read_csv()` and plot
> the at-home share against the threshold. That plot is a slide.

```{r}

```


```{r chunk-20}
dbDisconnect(con, shutdown = TRUE)
```

---

# Appendix — exercise answers

**0.** The `DESCRIBE` returns instantly because parquet stores schema in a footer at
the end of each file; DuckDB seeks to it and stops. No row data is touched.

**1a.** `SUM(CASE WHEN n_home_cbgs > 3 THEN 1 ELSE 0 END) / COUNT(*)`.

**1b.** `SELECT m.did, m.cbg FROM monthly m JOIN (SELECT did, MAX(dwell) AS mx FROM
monthly GROUP BY did) t ON m.did = t.did AND m.dwell = t.mx`. Worse because it reads
`monthly` twice, and because ties produce duplicate rows that then need deduplicating.

**1c.** Row counts and dwell will disagree for devices with many zero-dwell home rows
at one place and one long stay elsewhere. Trust the dwell version, but note that dwell
is a lower bound, so a place with many single-ping visits may be genuinely important
and look weightless.

**2a.** `9q8yy` decodes near 37.7°N, 122.4°W. `dhwux` should land near Fort Lauderdale,
matching the `city` field in the sample row.

**2b.** Rows without a centroid are ones whose block group isn't in the gazetteer —
usually water block groups or a vintage mismatch between the 2020 gazetteer and
whatever vintage Veraset used. A small share is fine; more than a percent means
checking vintages.

**2c.** No single right answer. Smaller thresholds use the geohash more often, which is
uniform but coarse; larger ones trust block groups that may be huge. The agreement
table tells you where block groups stop being more precise than 5 km.

**3a.** Suggested rule: flag devices where more than half their home dwell is more than
50 km from the assigned home block group, and exclude them. These are likely devices
with two residences, or ID recycling.

**3b.** `CREATE MACRO hav(lat1, lon1, lat2, lon2) AS 2 * 6371.0088 * asin(sqrt(
pow(sin(radians(lat2 - lat1) / 2), 2) + cos(radians(lat1)) * cos(radians(lat2)) *
pow(sin(radians(lon2 - lon1) / 2), 2)));`

**3c.** Only the speed, and probably not even that — DuckDB's optimiser pushes filters
down. Putting it earlier is clearer to a reader, which is a good enough reason.

**4a.** Home's share will move very little, because home dwell is in the hundreds of
minutes and τ is in the tens. The place τ matters is device-days with *no* home visit,
where it sets the relative weight of several short stops.

**4b.** `COALESCE(NULLIF(GREATEST(dwell, 0), 0), {TAU_MINUTES})`.

**4c.** A cap makes home comparable to a long stay rather than to a whole day, which
is closer to "where did this person spend discretionary time". It is a different
estimand and not the one Meta reports, so it belongs in the sensitivity section rather
than the headline.

**5a.** Grouping by `cbg` is right for a distance-based kernel — two POIs in one block
group are at the same distance, so they're the same observation for our purposes.
Grouping by `safegraph_place_id` would count them separately, which inflates trip
weighting for dense retail areas.

**5b.** `SELECT COUNT(*) FROM (SELECT did, day, cbg, loc_src, COUNT(DISTINCT d_km) AS
k FROM ... GROUP BY did, day, cbg, loc_src) WHERE k > 1` — safe here because it's one
day and a narrow group. A violation would mean the same block group resolved to two
different coordinates, i.e. a duplicated row in `cbg_lookup`.

**5c.** `banded |> group_by(did, band) |> summarise(share = sum(share) /
n_distinct(day), .groups = "drop")`.

**6a.** Expect the threshold to matter enormously in rural counties and little in
urban ones, because urban devices genuinely cluster within a block group of home.

**6b.** Rural counties should move most: a Wibaux device-day is home plus one long
trip, which is ~99% home by dwell and 50% by trip.

**6c.** In LA nearly all rows will be block-group-resolved and the sub-kilometre band
is meaningful. In Wibaux almost none will be, and it isn't.
